---
title: "Mean read counts and size"
date: "2026-07-14"
output:
    github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}

knitr::opts_chunk$set(message = F, warning = F)

library(kableExtra)
library(tidyverse)
library(ggtext)
library(ggh4x)
library(here)
library(knitr)

set_theme(theme_classic(base_size = 16) +
              theme(axis.text = element_text(size = 12)))

```

### Generates working dataframe

To work with tidy-compatible data, the .txt file generated from the scripts needs to be transformed into long format, in terms of metrics values, and then, again, converting it into wider format setting each metric name as columns.

Each sample is renamed to a reader-friendly format -- no one will understand our lab sample codes. Gene names were recoded, too. 

Finally, this work-friendly dataframe is converted into a .csv file for further analysis.

```{r}
df <- read.delim(here("data/raw_data/alignments/Stats/coverage.txt"),
                 check.names = F)

ampstats <- df %>%
    pivot_longer(cols = colnames(df[, 3:9]),
                 names_to = "GENE",
                 values_to = "VALUE") %>%
    pivot_wider(names_from = "METRIC", values_from = "VALUE")

# New relabeled name (remove lab ID)
sample_names <- paste("Sample", 1:9)

# Scientific-precise gene names
gene_names <- c("rpoB",
                "gyrA",
                "gyrB",
                "folP1",
                "folP2",
                "23S rRNA I",
                "23S rRNA II")

relabeled_ampstats <- ampstats %>%
    mutate(
        SAMPLE = factor(
            SAMPLE,
            levels = unique(ampstats$SAMPLE),
            labels = sample_names
        ),
        GENE = factor(GENE, levels = unique(ampstats$GENE), labels = gene_names)
    )

# Create a new directory to add all processed files
processed_files_path = "data/processed_data"
dir.create(here(processed_files_path), showWarnings = F)

write.csv(relabeled_ampstats,
          here(file.path(processed_files_path, "amplicon_stats.csv")),
          row.names = F)

glimpse(relabeled_ampstats)
```

### Total reads per gene

To understand how well each amplicon performed on this assay, we can sum up their number of reads and sort from the best to the worst.

```{r}

fill_genes <- c(
    "rpoB" = "#e41a1c",
    "gyrA" = "#ff7f00",
    "gyrB" = "#377eb8",
    "folP1" = "#4daf4a",
    "folP2" = "#a65628",
    "23S rRNA I" = "#984ea3",
    "23S rRNA II" = "#e6ab02"
)

relabeled_ampstats %>%
    group_by(GENE) %>%
    summarise(TOTAL_READS = sum(FREADS)) %>%
    ggplot(aes(
        x = TOTAL_READS,
        y = fct_reorder(GENE, TOTAL_READS),
        fill = GENE
    )) +
    labs(title = "Number of reads per gene", x = "Number of reads", y = NULL) +
    geom_col(width = 0.7, colour = "black", show.legend = F) +
    geom_text(aes(
        label = scales::label_number(
            accuracy = 1,
            scale_cut = scales::cut_short_scale()
        )(TOTAL_READS),
        hjust = 0,
        nudge_x = 1000
    )) +
    scale_x_continuous(
        breaks = seq(0, 70000, by = 10000),
        expand = expansion(mult = c(0, 0.15)),
        labels = scales::label_comma()
    ) +
    scale_fill_manual(values = fill_genes) +
    theme(axis.text.y = element_text(face = "italic"))

```

All amplicons generated more than 10K reads each. However, *rpoB*, and *folP2* generated the lowest reads counts (13K and 14K, respectively). Thus, it is neded to observe the results of other amplicon statistics (such as depth and distribution) to understand the performance of these two amplicons.

### Number of reads per sample

The number of reads is related to the coverage of each gene. Getting this count by samples tell about the quality of bacilli DNA after ExoV treatment (and also about the quality of our multiplex PCR).

```{r}

set.seed(20230407)

relabeled_ampstats %>%
    ggplot(aes(x = SAMPLE, y = FREADS)) +
    labs(
        title = "Number of reads per sample",
        x = NULL,
        y = bquote("Number of reads " (log[10] ~ "scale")),
        fill = NULL
    ) +
    geom_boxplot(
        outliers = F,
        colour = "black",
        width = 0.6,
        linewidth = 0.75
    ) +
    geom_jitter(
        aes(fill = GENE),
        size = 3,
        width = 0.1,
        shape = 21,
        alpha = 0.7
    ) +
    scale_y_continuous(
        breaks = 10 ^ (0:5),
        trans = "log10",
        limits = c(1, 10 ^ 5),
        label = scales::label_log()
    ) +
    scale_fill_manual(values = fill_genes) +
    coord_cartesian(clip = "off") +
    annotation_logticks(sides = "l", outside = T) +
    theme(
        legend.position = c(0.85, 0.25),
        legend.background = element_rect(fill = "transparent", colour = NA),
        legend.text.position = "left",
        legend.text = element_text(face = "italic"),
        axis.text.y = element_text(margin = margin(r = unit(10, "mm"))),
        axis.text.x = element_text(
            angle = 45,
            hjust = 1,
            vjust = 1
        )
    )

```

Sample 5 and 7 got the lowest number of reads. Maybe it is more informative to know the number of reads per gene in each sample Also, it could be nice to know the bacilli index of these two samples.

### Mean read size and coverage

Considering that we want to know the variants in these genes, one should know about the coverage of each amplicon that our sequencing could recover.

```{r}

ampsize <- c(
  "rpoB"        = 469, 
  "gyrA"        = 575, 
  "gyrB"        = 520, 
  "folP1"       = 797, 
  "folP2"       = 757, 
  "23S rRNA I"  = 405, 
  "23S rRNA II" = 565
)

relabeled_ampstats %>%
    mutate(AMPSIZE = ampsize[as.character(GENE)], COVERAGE = READSIZE / AMPSIZE) %>%
    group_by(GENE, AMPSIZE) %>%
    summarise(
        MEAN_READSIZE = mean(READSIZE),
        SD_READSIZE = sd(READSIZE),
        MEAN_COVERAGE = mean(COVERAGE) * 100,
        SD_COVERAGE = sd(COVERAGE) * 100,
        .groups = "drop"
    ) %>%
    mutate(
        READSIZE_LABEL = sprintf("%.1f ± %.1f", MEAN_READSIZE, SD_READSIZE),
        COVERAGE_LABEL = sprintf("%.1f ± %.1f", MEAN_COVERAGE, SD_COVERAGE)
    ) %>%
    select(GENE, AMPSIZE, READSIZE_LABEL, COVERAGE_LABEL) %>%
    rename(
        Gene = GENE,
        `Amplicon Length (bp)` = AMPSIZE,
        `Mean read length (bp)` = READSIZE_LABEL,
        `Mean read coverage (%)` = COVERAGE_LABEL
    ) %>%
    kable(format = "markdown") %>%
    kable_classic() %>%
    column_spec(1, italic = T)
```

Overall, more than 70% of all amplicons -- except *folP1* -- could be recovered. Maybe, *folP1* mean read size was biased by its number of reads in samples 2 and 5. In both, *folP1* was the least sequenced gene. Again, the bacilli index of these samples can provide interesting insights. Could *M. leprae* genome be degraded in low bacillary index samples?