---
title: "Untitled"
date: "2026-07-14"
output:
    github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}

knitr::opts_chunk$set(
    message = F,
    warning = F
)

library(kableExtra)
library(tidyverse)
library(ggtext)
library(ggh4x)
library(here)
library(knitr)

set_theme(
    theme_classic(base_size = 16) +
        theme(axis.text = element_text(size = 14))
)

```

### Generates 

```{r}
df <- read.delim(here("data/raw_data/alignments/Stats/coverage.txt"),
                 check.names = F)

ampstats <- df %>%
    pivot_longer(cols = colnames(df[, 3:9]),
                 names_to = "GENE",
                 values_to = "VALUE") %>%
    pivot_wider(names_from = "METRIC", values_from = "VALUE")

# New relabeled name (remove lab ID)
sample_names <- paste("Sample", 1:9)

# Scientific-precise gene names
gene_names <- c("rpoB",
                "gyrA",
                "gyrB",
                "folP1",
                "folP2",
                "23S rRNA I",
                "23S rRNA II")

relabeled_ampstats <- ampstats %>%
    mutate(
        SAMPLE = factor(
            SAMPLE,
            levels = unique(ampstats$SAMPLE),
            labels = sample_names
        ),
        GENE = factor(
            GENE, 
            levels = unique(ampstats$GENE), 
            labels = gene_names
        )
    )

glimpse(relabeled_ampstats)

# Create a new directory to add all processed files
processed_files_path = "data/processed_data"
dir.create(here(processed_files_path), showWarnings = F)

write.csv(relabeled_ampstats,
          here(file.path(processed_files_path, "amplicon_stats.csv")),
          row.names = F)
```

### Total reads per gene

To understand how well each amplicon performed on this assay, we can sum up their number of reads from all samples and sort from the best to the worst.

```{r}

fill_genes <- c(
    "rpoB" = "#e41a1c",
    "gyrA" = "#ff7f00",
    "gyrB" = "#377eb8",
    "folP1" = "#4daf4a",
    "folP2" = "#a65628",
    "23S rRNA I" = "#984ea3",
    "23S rRNA II" = "#e6ab02"
)

relabeled_ampstats %>%
    group_by(GENE) %>%
    summarise(TOTAL_READS = sum(FREADS)) %>%
    ggplot(aes(
        x = TOTAL_READS,
        y = fct_reorder(GENE, TOTAL_READS),
        fill = GENE
    )) +
    labs(title = "Number of reads per gene", x = "Number of reads", y = NULL) +
    geom_col(colour = "black", show.legend = F) +
    geom_text(aes(
        label = scales::label_number(
            accuracy = 1,
            scale_cut = scales::cut_short_scale()
        )(TOTAL_READS),
        hjust = 0,
        nudge_x = 1000
    )) +
    scale_x_continuous(
        breaks = seq(0, 70000, by = 10000),
        expand = expansion(mult = c(0, 0.15)),
        labels = scales::label_comma()
    ) +
    scale_fill_manual(values = fill_genes) +
    theme(axis.text.y = element_text(face = "italic"))

```

In general, all amplicons generated more than 10K reads each. However, rpoB, and folP2 generated the fewer number of reads (13K and 14K, respectively). In this case, we need to observe the results of other amplicon statistics to infer about the sequencing quality of these two amplicons.

### Number of reads per sample

The number of reads is related to the coverage of each gene. Getting this count by samples tell about the quality of bacilli DNA after ExoV treatment.

```{r}

set.seed(20230407)

relabeled_ampstats %>%
    ggplot(aes(x = SAMPLE, y = FREADS)) +
    labs(
        title = "Number of reads per sample",
        x = NULL,
        y = bquote("Number of reads " (log[10] ~ "scale")),
        fill = NULL
    ) +
    geom_boxplot(
        outliers = F,
        colour = "black",
        width = 0.6,
        linewidth = 0.75
    ) +
    geom_jitter(
        aes(fill = GENE),
        size = 3,
        width = 0.1,
        shape = 21,
        alpha = 0.7
    ) +
    scale_y_continuous(
        breaks = 10 ^ (0:5),
        trans = "log10",
        limits = c(1, 10 ^ 5),
        label = scales::label_log()
    ) +
    scale_fill_manual(values = fill_genes) +
    coord_cartesian(clip = "off") +
    annotation_logticks(sides = "l", outside = T) +
    theme(
        legend.position = c(0.85, 0.25),
        legend.text.position = "left",
        legend.text = element_text(face = "italic"),
        axis.text.y = element_text(margin = margin(r = unit(10, "mm"))),
        axis.text.x = element_text(
            angle = 45,
            hjust = 1,
            vjust = 1
        )
    )

```

Sample 5 and 7 got the lowest number of reads. Maybe it is more informative to know the number of reads per gene in each sample

```{r}

ampsize <- c(
  "rpoB"        = 469, 
  "gyrA"        = 575, 
  "gyrB"        = 520, 
  "folP1"       = 797, 
  "folP2"       = 757, 
  "23S rRNA I"  = 405, 
  "23S rRNA II" = 565
)

relabeled_ampstats %>%
    mutate(AMPSIZE = ampsize[as.character(GENE)], COVERAGE = READSIZE / AMPSIZE) %>%
    group_by(GENE, AMPSIZE) %>%
    summarise(
        MEAN_READSIZE = mean(READSIZE),
        SD_READSIZE = sd(READSIZE),
        MEAN_COVERAGE = mean(COVERAGE) * 100,
        SD_COVERAGE = sd(COVERAGE) * 100
    ) %>%
    mutate(
        READSIZE_LABEL = sprintf("%.1f ± %.1f", MEAN_READSIZE, SD_READSIZE),
        COVERAGE_LABEL = sprintf("%.1f ± %.1f", MEAN_COVERAGE, SD_COVERAGE)
    ) %>%
    select(GENE, AMPSIZE, READSIZE_LABEL, COVERAGE_LABEL) %>%
    rename(
        Gene = GENE,
        `Amplicon Length (bp)` = AMPSIZE,
        `Mean read length (bp)` = READSIZE_LABEL,
        `Mean read coverage (%)` = COVERAGE_LABEL
    ) %>%
    kable(format = "markdown") %>%
    kable_classic() %>%
    column_spec(1, italic = T)
```

