---
title: "Proportion of aligned and unaligned reads"
date: "2026-07-15"
output:
    github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}

knitr::opts_chunk$set(message = F, warning = F)

library(kableExtra)
library(scales)
library(tidyverse)
library(knitr)
library(here)

set_theme(theme_classic(base_size = 16) +
              theme(axis.text = element_text(size = 12)))

```

### Number of aligned and unaligned reads

```{r}
df <- read.delim(here("data/raw_data/alignments/Stats/alignment.txt"),
                 check.names = F)

glimpse(df)
```

### Distribution of aligned and unaligned reads

The proportion of reads that align to the amplicon sequences can tell about how well we were able to recover the sequences of each amplicon.

In this sense, these values will tell us if we were able to get enough reads from slit skin smears.

```{r}
relabeled_alnstats <- df %>%
    mutate(SAMPLE = factor(
        SAMPLE,
        levels = unique(df$SAMPLE),
        labels = paste("Sample", 1:9)
    ))

relabeled_alnstats %>%
    mutate(
        prop_aligned = ALIGNED / TOTAL * 100,
        prop_unaligned = UNALIGNED / TOTAL * 100,
        prop_f = FORWARD / ALIGNED * 100,
        prop_r = REVERSE / ALIGNED * 100,
        prop_both = BOTH / ALIGNED * 100
    ) %>%
    mutate(
        aln_label = sprintf("%d (%.1f%%)", ALIGNED, prop_aligned),
        unaln_label = sprintf("%d (%.1f%%)", UNALIGNED, prop_unaligned),
        f_label = sprintf("%d (%.1f%%)", FORWARD, prop_f),
        r_label = sprintf("%d (%.1f%%)", REVERSE, prop_r),
        both_label = sprintf("%d (%.1f%%)", BOTH, prop_both),
        total_recoded = label_number(
            scale = 1e-3,
            accuracy = 1,
            suffix = "k"
        )(ALIGNED)
    ) %>%
    select(SAMPLE,
           total_recoded,
           aln_label,
           unaln_label,
           both_label,
           f_label,
           r_label) %>%
    rename(
        Sample = SAMPLE,
        Total = total_recoded,
        Aligned = aln_label,
        `Aligned with both primers` = both_label,
        `Aligned with forward` = f_label,
        `Aligned with reverse` = r_label,
        Unaligned = unaln_label
    ) %>% 
    kable() %>% 
    kable_classic()
```

Except for samples 2 and 5, more than 50% of the reads were aligned to the target sequences in all samples. The majority of the reads had both forward and reverse primers. However, only in 7 and 9 more than 70% of the reads had both primers.

### Proportion of aligned x unaligned

Just a better visualisation of the proportion of aligned and unaligned reads per sample.

```{r}
relabeled_alnstats %>%
    mutate(aligned_prop = ALIGNED / TOTAL) %>%
    select(SAMPLE, UNALIGNED, ALIGNED, aligned_prop) %>%
    pivot_longer(
        cols = c(UNALIGNED, ALIGNED),
        names_to = "reads",
        values_to = "number"
    ) %>%
    ggplot(aes(
        x = fct_reorder(SAMPLE, -aligned_prop),
        y = number,
        fill = reads
    )) +
    labs(x = NULL, y = "Aligned proportion", fill = NULL) +
    geom_col(width = 0.7, position = position_fill(reverse = T)) +
    scale_y_continuous(expand = expansion(), labels = label_percent()) +
    scale_fill_manual(labels = c("Aligned", "Unaligned"), values = c("#2166ac", "#e0e0e0")) +
    theme(legend.position = "top",
          axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

Notably, the two samples with the lowest number of reads had the lowest proportion of aligned reads. Again, this can be related to the bacillary index of those samples. Even so, we were able to sequence a significative amount of sequences in all samples.